Transitioning Towards Continuous Delivery in the B2B Domain: A Case Study

### 2015-03-05 paper.tex
Version 1.0
Updated email
Final

% This is lnbip.tex the demonstration file of the LaTeX macro package for
% Lecture Notes in Business Information Processing from Springer-Verlag.
% It serves as a template for authors as well.
% version 1.0 for LaTeX2e
%
\documentclass[lnbip]{svmultln}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{graphicx}
%\usepackage{graphics} 
% \makeindex          % be prepared for an author index

@@

\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quotation}}
  {\signed{\usebox\mybox}\end{quotation}}
%
\begin{document}
%
\mainmatter              % start of the contribution
%
\title{Transitioning Towards Continuous Delivery in the B2B Domain: A Case Study}
%
%TODO: check this
\titlerunning{Continuous Delivery in B2B Domain}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Olli Rissanen \inst{1,2}, J{\"u}rgen M{\"u}nch \inst{1}
%Jeffrey Dean \and David Grove \and Craig Chambers \and Kim~B.~Bruce \and
}
%TODO: check this
\authorrunning{Olli Rissanen}   % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Olli Rissanen}
%
\institute{Department of Computer Science, University of Helsinki, \\
P.O. Box 68, FI-00014 University of Helsinki, Finland
\and
Steeri Oy, Tammasaarenkatu 5, 00180 Helsinki, Finland
\email{olli.rissanen@steeri.fi, juergen.muench@cs.helsinki.fi}
}

@@

\begin{abstract}        % give a summary of your study
Delivering value to customers in real-time requires companies to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction and helps to ensure that the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article presents a case study that is conducted in a medium-sized software company operating in the B2B domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. The company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers require to perform manual acceptance testing, while some are reluctant towards new versions. By utilizing continuous delivery, it is possible for the case company to shorten the feedback cycles, increase the reliability of new versions, and reduce the amount of resources required for deploying and testing new releases.

%s such as long feedback cycles, low reliability in new versions and high amount of resources required for releases.

%                         please supply keywords within your abstract
\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}
To deliver value fast and to cope with the increasingly active business environment, companies have to find solutions that improve efficiency and speed. Agile practices \cite{cockburn2002agile} have increased the ability of software companies to cope with changing customer requirements and changing market needs \cite{dzamashvili2010impact}. To even further increase the efficiency, shortening the feedback cycle enables faster customer feedback. Continuous delivery is a design practice that aims to shorten the delivery cycles by automating the deployment process and frequently deploying the software to customer environment. In continuous delivery, the software is developed in a way that it is always ready for releasing. %While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, continuous delivery adds automated acceptance testing and automated deployment.

This study is an exploratory case study, which explores how continuous delivery can be applied in the case company that operates in the B2B domain. While existing studies of applying the design practice exist \cite{olsson2012climbing, neely2013continuous}, none of the studies focuses specifically in the B2B domain. This study specifically aims to identify the main requirements, problems and key success factors with regards to continuous delivery in this domain. Extending the development process towards continuous delivery requires a deep analysis of the current development and deployment process, seeking the current problems and strengths. Adopting continuous delivery also requires understanding the requirements of continuous delivery, and restrictions caused by the developed software products.

@@

\section{Background and Related Work}
In an agile process software release is done in periodic intervals \cite{cockburn2002agile}. Compared to waterfall model it introduces multiple releases throughout the development. Continuous delivery, on the other hand, attempts to keep the software ready for release at all times during development process \cite{cdbook}. Continuous delivery is an extension to continuous integration, where the software functionality is deployed frequently at customer environment. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error, documents required for the build and increases confidence that the build works \cite{cdbook}. It therefore aims to solve the problem of how to deliver an idea to users as quickly as possible.

%release cycles
 %Instead of stopping the development process and creating a build as in an agile process, the software is continuously deployed to customer environment. This doesn't mean that the development cycles in continuous delivery are shorter, but that the development is done in a way that makes the software always ready for release.

%\begin{figure}
% use the \includegraphics command from the graphics package
% (N.B. put \usepackage{graphics} into your preamble then)
% for inclusion of your images. A sample call could read:
% \includegraphics{myimage}
%\vspace{2.5cm}
%\caption{This is the caption of the figure displaying a white eagle and
%a white horse on a snow field}
%\end{figure}
%delivery vs deployment
Continuous delivery differs from continuous deployment. Both continuous delivery and continuous deployment include automatic deployment to a staging environment. Continuous deployment includes deployment to a production environment, while in continuous delivery the deployment to a production environment is done manually. The purpose of continuous delivery is to prove that every build is deployable \cite{cdbook}. An essential part of continuous delivery is the deployment pipeline, which is an automated implementation of an application’s build, deploy, test and release process \cite{humble2006deployment}. A deployment pipeline can be loosely defined as a consecutively executed set of validations that a software has to pass before it can be released. Common components of the deployment pipeline are a version control system and an automated test suite. Humble and Farley define the deployment pipeline as a set of stages, which cover the path from the commit stage to a build.

%Barriers in CI->CD transition: stairway
Challenges in adopting continuous delivery have been researched in multiple studies. Olsson et al. investigate the organization evolution path and the transition phase from continuous integration to continuous delivery \cite{olsson2012climbing}.  The authors define continuous delivery as one of the final steps in the organization evolution path. The authors identify barriers that companies need to overcome to achieve the transition. One such barrier is the custom configuration at customer sites. Maintaining customized solutions and local configurations alongside the standard configurations creates issues. The second barrier is the internal verification loop, that has to be shortened not only to develop features faster but also to deploy fast. Finally, the lack of transparency and getting an overview of the status of development projects is seen as a barrier.

@@

%The practice of deploying versions continuously allows for faster customer feedback, ability to learn from customer usage data and to focus on features that produce real value for the customer \cite{olsson2012climbing}. Olsson et al. state that in order to shorten the internal verification loop, different parts in the organization should also be involved, especially the product management as they are the interface to the customer \cite{olsson2012climbing}. They also note that finding a pro-active lead customer is essential to explore and form a new engagement model.

@@

%The purpose of this study is to seek ways to improve the product development process of two software products of the case company in the B2B domain by utilizing continuous delivery. This includes analysing the challenges and benefits of this practice. Existing documented implementations of continuous delivery are primarily executed in the B2C domain, often with a Software as a Service (SaaS) product. The focus of this study is in the B2B domain, with two software products that are not used as SaaS and are developed directly to other companies acting as customers. The characteristics of this domain are typically longer release cycles and customer owned environments.

@@

\noindent Software development practices and product characteristics vary based on the domain and delivery model. Typical B2C applications are hosted as Software as a Service (SaaS) applications, and accessed by users via a web browser. In the B2B domain, applications installed to customer environments are very common. The purpose of this research question is to identify the challenges faced in applying continuous delivery in the B2B environment. The research question is answered by conducting interviews to discover the current development process and its challenges in the case company, and using these findings and the available literature on continuous delivery to map the initial set of challenges these approaches will encounter in the case company. The available literature is used to provide a thorough understanding of continuous delivery as a whole, so that challenges can be identified in all aspects of the practice.
%Question: contractual issues? 
%Question: selecting the correct OEC in b2b is harder. for example revenue. 
%question: in B2C people develop their own product, but in B2B the feature ideas might come only from customer.
\newline

%collect ideas from literature review and interview
%\noindent \textbf{RQ 2: How to collect usage data and customer feedback in B2B domain, where user amounts can be very low?}

%\noindent Research question 4 is answered by analysing the usage data and customer feedback collection of the case company. This is carried out by interviewing the key persons responsible for customer interaction and developers responsible for usage data collection.

%As compared to the B2C domain where the product is often deployed in cloud with a lot of users, B2B applications might only have a handful of users. This question %is answered by collecting ideas from the literature review and interview results. \newline
%-Legal issues
%-Technical issues

%Interview results have to be compared to literature review results
\noindent \textbf{RQ2: How does continuous delivery benefit the case company?} %What are the problems and strengths of the current development process?

@@

\subsubsection{Research design}
In this research, the units under the study are two teams within the case company, and the two software products developed by these teams. By focusing on two different products, a broader view on the application and consequences of the development approach can be gained. The first product, a marketing automation, is used through an extensive user interface. The second product under inspection is a Master Data Management \cite{loshin2010master} solution running as an integrated background application.

%Methods
The primary source of information in this research are semi-structured interviews \cite{runeson2009guidelines}, performed within the two teams under study. The interview consists of pre-defined themes focusing on current development process, current deployment process, current interaction with customers, the software products and future ways with continuous delivery. Data is also collected through the product description documents and development process documents to verifying and supplementing the interview data. Data triangulation is implemented by interviewing multiple individuals in different positions of the teams. Methodological triangulation is implemented by collecting documentary data regarding the development process from internal documents, and by including observations made by the author in the analysis.

@@

\section{Results}

%When to use CD: http://www.continuousagile.com/unblock/cd_costs_benefits.html
%You provide an online service: Web, SaaS, PaaS, online gaming, online big data, or mobile apps with Web service back ends. If you provide online services, you need to move to continuous delivery, or your competitors will race ahead of you.
This section is structured according to the research questions. The challenges regarding continuous delivery are analyzed in three areas: technical, procedural and customer. Technical aspect includes the environmental challenges, configural challenges and other challenges related to the software product and its usage. Procedural aspect includes the challenges regarding the development process of software. Customer aspect consists of the customer interaction process and customer commitment.

@@

%-testing costs could be pushed at least partially to users
The developers perceive automated testing and test environments to be the largest technical task. The developers state that building a sufficient test automation is a very laborious process especially due to the maturity of the software, and are concerned with the maintainability of the test suite. The management is not sure what to test with automatic acceptance testing to validate that a version is valid.

@@

%External components
It is also common for B2B applications to have external components that have to be configured when the software is installed or the APIs to these components changed. The configurations for these external components either have to be manually updated, or automated as well. One of the main differences between B2B and B2C domains is the production environment. Both of the case company's products are used in multiple different customer environments. This introduces a problem of managing different configurations per customer environment and software instance.
%Different configuration per customer

 %This introduces a problem of managing different configurations per customer environment and software instance. While companies operating in the B2C domain often own the production environment, in B2B domain it is common to install the software on customer environment. The customer owned environment introduces multiple challenges: accessing the environment, transferring the software product to the environment and integrating the software to other components within the environment. The challenges related to transferring the software include firewall configurations, required certifications and user rights. The customers also have to be informed on actions performed in their servers. For certain environments, a VPN connections is required. A VPN connection often has a user limit, which can prevent automatic version deployment.

%According to Jan Bosch, "In an IES environment, there is only one version: the currently deployed one. All other versions have been retired and play no role." \cite{bosch2012building}".

%Database versioning
%database needs to be versioned

%Firewall configurations
%Required certifications
%Missing user rights
%VPN user limit
%Environment

%-automate testing
%Continuous delivery requires you to set up and administer automated testing. You will need to write automated tests, and you will need to automate the configuration of test environments. In our experience, automating testing takes more time and effort than any other aspect of a continuous delivery rollout.

%-testing costs could be pushed at least partially to users
%In an experimental product, you can push some of the costs of testing onto users. We'll talk about this option when we cover testing and product management.

%Tällä hetkellä tiimin resurssit eivät riitä sen toteutukseen. Asiakkaiden pitää ymmärtää mistä on kyse. Asiakkailta pitää saada tähän lupa / sopimus. Tällä hetkellä tekninen haaste eniten on tietokannan versiointi johon on olemassa ratkaisuja. Integraatioiden automaattinen päivittäminen vaatisi ehkä kohtuuttoman paljon resursseja suhteessa siihen mitä saavutetaan. Itse dialogin päivitysten automatisointi olisi tehtävissä helpoiten. Yleinen smoketestaus olisi tehtävissä, mutta vaatii että testit kirjoitetaan, pitäisi tehdä paljon, ja niitä pitäisi ylläpitää. Tällä hetkellä tähän ei ole resursseja.  Ei voida siirtää aina uutta versiota, koska pitää pystyä takaamaan isolla varmuudella että asiat toimivat. Käytännössä se että sovitaan asiakkaan kanssa että vaihdetaan versio, ja tällöin laukastausiin automaattinen tuotantoonsiirto, olisi toimiva ratkaisu.

%External components have to be configured
%Useampi asia joka pitää laittaa käyttöön, OpenAM, DJ jne. Jos yksikin unohtuu laittaa käyntiin, jotkin tuotteen palvelut eivät toimi. (Ulkoiset komponentit) SaaS-tuotteena herokussa ei ole ongelmaa.

%Database versioning

%Different configurations per customer

%Firewall configuratiosn

%In the IT-departments, customers have to be informed on actions in their servers. There might be firewall configurations, required certifications and missing %rights. 
%Required certifications
%Missing rights
%VPN user limit
%siirto voi olla haaste, sillä ympäristöt ovat asiakkaiden hallussa. yhteyden saaminen esim. windows-koneille vaatii VPN:n, eikä se ole aina yksinkertaista. esimerkiksi VPN:n käyttäjälimitit asiakkaille voivat olla täynnä.

%Integrating the software to third-party components

%Downtime

  %IT-puolelle täytyy ehkä ilmoitella mitä tapahtuu, tai pyytää heitä tekemään asioita mihin meillä ei ole oikeuksia. Saattaa olla palomuurikonffeja joita pitää säätää, tai toimittaa meille jotain sertifikaatteja.

%(CDM) -Downtime -ei ihmiskäyttäjiä, olisi ihan tehtävissä. UAT:kin on, jengi varaa resurssit testaamiseen, en tiedä onko kykyä tehdä kaksi versiota ja deployata ne tuotantoon. jos asiakkaan testaajien pitää kerran viikossa mennä UAT:hen testaamaan uutta versiota, söisi se niiltä paljon resursseja. Ei uusi versio kuitenkaan huono tilanne ole, ei heidän välttämättä tarvitsisi sitä testata. UAT on käytännössä pakollinen ennen tuotantoa, Sanomilla on sitä testattu ja se on CDM:n kanssa varsin haastavaa kun pitäisi tietää mitä tulisi eksakstisti takaisin. pitää pelata ulkositen palvelijoiden (VTJ) pelisäännöillä kun ei voida tietää mitä palautuu. yhdellä datasetillä testaaminen on vähän turhaa, kun yhden datasetin toimiminen ei kerro meille käytännössä mitään.  Ei tietoa teknisistä haasteista.

@@

%UAT
%-UAT is often the norm and is required before anything is moved to production
%-in B2B, continuous deployment straight to production is very risky
The basic deployment pipeline in the case company first includes a deploy to a user acceptance testing server, which is then tested manually by either the team or the customer. Only after the version has been acceptance tested and validated to work properly, can the production version be released. Continuous deployment to production is seen very risky due to the applications playing a major role in running the customers business. %However, continuous deployment to a staging environment is seen as very beneficial. %Therefore the deployment procedure has to be designed to allow testing the software in a user acceptance testing environment, and making it possible to manually trigger the deployment to production.

Both of the case company's products are developed with a branching model, where feature branches are first thoroughly developed and then integrated to the master branch. With continuous delivery the long-lived feature branches should be changed to short-lived and relatively small feature branches to allow exposing new functionality faster to the customers, and receive feedback faster. While the small feature branches might be common for companies with a relatively new software products, companies that have been developing products for a long time might be more devoted to the practice of feature branches. %However, it is still possible to keep the software releasable with long-lived feature branches by hiding new functionality or by constructing it as a series of releasable small changes \cite{cdbook}.
%If the components are sufficiently small, it is preferable to have a single build for the whole application—giving the fastest feedback of all. \cite{cdbook}

%The benefit of smaller branches is that while there may still be bugs, they're introduced in a fewer number of commits, and entire large features don't have to be rollbacked. Smaller branches also enable faster feedback.

%Pull requests
%Triggering
%
The software applications in B2B often are large and modular applications, as is the case in the case company. The point when a deployment is triggered has to be designed to maintain the integrity of the application. As the deployment process is currently manually triggered by first releasing a version, a suitable time can be chosen each time. With continuous delivery the deployment process is triggered automatically, and with a modular project each module has to be in the correct state in order to produce a coherent version.
%\begin{aquote}{Developer}
%"There are dependencies between projects, and occasionally merges are done without the will to deploy. However, if pull requests are to be deployed when merged, a better picture of this process is needed for the developers."
%\end{aquote}
%In the case company, merging a pull request is not a valid trigger of a deployment, since pull requests are project specific and the software products are composed of multiple projects. Some features require changes to multiple projects, and all of the pull requests have to be merged for the feature to work as intended. However, a good CI tool will handle the dependency management \cite{cdbook}.

%Multiple customer environments
Both of the case company's products are used by multiple customers, each having their own environments. As the deployments are currently done manually, the customers receiving each deployment can be manually chosen.However, with a continuous delivery process whenever a feature or a new release is ready to be delivered, it can either be deployed to a single customer or to every customer. %Attention has to be paid when configuring the workflow to allow deploying to an arbitrary customer. If only certain customers accept updates on a continuous basis, it has to be possible to prohibit deploying to every customer at once. %This is the case especially in the early stages, where all of the customers projects aren't yet in the state of continuous delivery.

%Versioning
Multiple customer environments affects versioning of the software product. In the case company, each customer has a unique configuration of the product, with possibly different versions of certain components. According to Jan Bosch, in an Innovation Experiment System environment only a single version exists: the currently deployed one. Other versions are retired and play no role \cite{bosch2012building}. However, with multiple environments, multiple different versions of the software are necessary at least in the early phase. %Eventually the product may be developed in such manner that a single version is sufficient. The build script therefore has to be configured to consider the customer-specific components when compiling the builds for each customer.

%"A final aspect is the avoidance of versioning of software. Traditionally, for many companies, every customer would ultimately have a unique configuration of the product with different versions of components making up the system. This adds a whole new layer of complexity to the already costly process of deploying new versions. In an IES environment, there is only one version: the currently deployed one. All other versions have been retired and play no role."

@@

Due to increased developer responsibility and varying interval of version updates continuous delivery causes, a team leader expresses concern that the delivery process complicates tracking when deployments are performed, and when features are finished. This also concerns other parties working in the customer interface, such as sales. %While this is a concern for version deployments to the UAT environment, especially if developers perform these deployments at will, can production releases still be negotiated with the customer. Neely and Stolt have also found similar results \cite{neely2013continuous}, which they solved by providing mechanisms by which the parties if need of knowledge regarding release times could track the work in real time.

%Neely and Stolt found out that with a continuous flow, Sales and Marketing departments lost the track of when features are released \cite{neely2013continuous}. Continuous delivery requires the company as a whole to understand the continuous flow process. Neely and Stolt solved this challenge by providing mechanisms by which the parties in need of knowledge regarding release times could track the work in real time \cite{neely2013continuous}.

%-should the product be always deployed to every customer?

%jos pull requestit haluttaisiin deployata automaattisesti, eivät pullrequestit ole aina täysin toimivia sellaisenaan. riippuvuuksia niidän välissä, ja toisinaan mergetään ilman että halutaan deployauksia. nykyään tämä tieto liikkuu kehittäjien päässä. pitäisi saada tarkempi kuva jos pullrequestin triggeröinti aiheuttaisi deployauksen. yksi haaste on se että asiakkaita ja ympäristöjä on paljon. miten määritetään mille asiakkaalle deployataan? ei ole järkeä viedä asiakkaille jotka eivät ole sitä toivoneet. täysin automaattisesti vaikea tehdä, ja ihmisen pitäisi päättää mihin versio deployataan. muutokset saattavat vaatia muiden komponenttien päivityksen, tietokanta jne. niiden koordinointi voi olla haaste. myös jos CDM on integroitu muiden asiakkaiden tuotteisiin saattaa rajapinta vaihtua. siirto voi olla haaste, sillä ympäristöt ovat asiakkaiden hallussa. yhteyden saaminen esim. windows-koneille vaatii VPN:n, eikä se ole aina yksinkertaista. esimerkiksi VPN:n käyttäjälimitit asiakkaille voivat olla täynnä.

%CD costs http://www.continuousagile.com/unblock/cd_costs_benefits.html

%Costs

%-developer training: social commitment and understaind of the version control system
%Continuous delivery also involves developer training costs. Developers need to understand the details of their version control system. They need a full social commitment to automated testing. They must take more responsibility for delivering release-ready code and features.

%-understanding how major changes can be inserted and hidden unless they're ready
%Incremental releases are complicated. Developers must figure out how to insert major changes invisibly, and hide them with switches until they are ready to be released. This staged migration adds some complexity that is not present if you just rip out the old stuff and then test the new stuff. However, you will gain confidence by seeing changes working in production systems before you "unveil" them to your users. This can more than compensate for the increased complexity
%It is faster. You unblock your team so that developers can work at full speed. They don't have to wait for a product manager, product owner, or marketing person to tell them that users, marketing and documentation are ready for a release. Instead, they release everything that is ready, and hide it from the general public. The PM or marketing person can come along later, see that the new feature works, and start the "unveil" process.
%It provides the correct incentives. The developer is responsible for the release, because he or she is the one who can fix it. A developer can be called back from Friday night beers to fix a problem. Therefore, developers have a compelling reason to make a good decision about what to release. They should not throw buggy code over the all to a QA person. They need to be forced to focus on quality by removing the QA training wheels. They will be motivated to build good tests, and good scripts for testing and deployment.

%Developer responsibility
%Developers gain a lot of power and a lot of responsibility from continuous delivery.
%Developers (programmers) will decide what is ready to release. This is a strong finding. We see it in all successful continuous delivery organizations. The developer of a feature decides when it is ready, and runs the scripts to put it into the production version.

%http://www.slideshare.net/wakaleo/continuous-deliverywithmaven CD with maven

%http://blogs.atlassian.com/2014/07/skeptics-guide-continuous-delivery-part-3-real-world-pipelines/ Pipeline

%en näe että sitä saataisiin ikinä toimimaan nykyisellä kokoonpanolla. cdm koostuu niin monesta projektista, että käytännössä en tiedä miten voisimme automatisoida se. riippuen mihin projektiin on tuulut muutoksia, pitää ne releaseta ja laittaa cdm:n pääprojektiin joka ottaa muuttuneet releaset mukaan. release-numerot vaihtelee per aliprojekti, voi olla joko pieni- tai iso releasemuutos ja sen mukaan kasvatetaan releasenumeroa. jos olisi vain yksi projekti niin voitaisiin se automatisoida. yhden projektin tekemisessä ei sinäänsä ole haasteita. monen projektin idea on tällä hetkellä aika huono, sillä kun buildaamme cdm:n, buildaamme aina pääprojektin. yksikään aliprojektia ei buildata koskaan, pl. käyttöliittymä. hyöty siitä että ne on itsenäisiä projekteja on aika nolla. nyt kaikissa aliprojekteissa omat versionumerot, joka on aika huono ratkaisu. jos kaikki aliprojektit olisi kiinnitetty samaan versionumeroon koko CDM:llä, voitaisiin vain päättää että nyt ollaan tehty tällaiset muutokset CDM:ään. mietitään että meillä on githubissa kamat, jos haluttaisiin autodeployaa tuotantoversio asiakkaille, saataisiinko esimerkiksi pääsy kaikkiin asiaksympäristöihin että voisimme automaattisesti siirtää eri releasen? voisiko kantamigraatiot automatisoitua? mitä jos tuotantodataa pitäisi päivittää, saataisiinko sitä päivitettyä, tuskin.

%UAT

@@

%Customers have been trained to use a certain version of the product
Some customers of the case company are reluctant towards new releases. One of the reasons for this reluctancy is that new releases occasionally contain new bugs. In the case company, customers have been trained to perform certain tasks with a certain user interface. The customer might perform these tasks daily, once every two weeks or even less frequently. If the UI changes often, the customers feel lost and initially take more time to perform the tasks. This causes frustration in the users, and visible changes generally increases the reluctancy customers have towards new versions, unless the changes are significantly improving the user experience.  
\begin{aquote}{Product owner}
"The user interface should be easy to use. Now it's relatively hard to learn. If customers have just learned to perform a task, and we change the UI, the feedback is terrible."
\end{aquote}

%Customer reluctancy towards new versions
%While user experience could be improved in the long run, customers might feel frustrated when they have to learn to work with modifications. If modifications were made more often, it leads to smaller and less notable modifications which might reduce the customer reluctancy towards new versions. Especially UI changes cause frustration.
%\begin{aquote}{Product owner}
%"If something changes, communication with the customer is hard. Even if the customers felt that the version included very good changes, if something visible changes it gets alot of negative feedback".
%\end{aquote}

>>>>

% This is lnbip.tex the demonstration file of the LaTeX macro package for
% Lecture Notes in Business Information Processing from Springer-Verlag.
% It serves as a template for authors as well.
% version 1.0 for LaTeX2e
%
\documentclass[lnbip]{svmultln}
\usepackage{graphicx}
\usepackage{makeidx}  % allows for indexgeneration
%\usepackage{graphics} 
% \makeindex          % be prepared for an author index

@@

\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quotation}}
  {\signed{\usebox\mybox}\end{quotation}}
%
\begin{document}
%
\mainmatter              % start of the contribution
%
\title{Transitioning Towards Continuous Delivery in the B2B Domain: A Case Study}
%
\titlerunning{Continuous Delivery in the B2B Domain}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Olli Rissanen \inst{1,2}, J{\"u}rgen M{\"u}nch \inst{1}
}
\authorrunning{Olli Rissanen}   % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Olli Rissanen}
%
\institute{Department of Computer Science, University of Helsinki, \\
P.O. Box 68, FI-00014 University of Helsinki, Finland
\and
Steeri Oy, Tammasaarenkatu 5, 00180 Helsinki, Finland
\email{olli.rissanen@aalto.fi, juergen.muench@cs.helsinki.fi}
}

@@

\begin{abstract}        
Delivering value to customers in real-time requires companies to utilize real-time deployment of software to expose features to users faster, and to shorten the feedback loop. This allows for faster reaction and helps to ensure that the development is focused on features providing real value. Continuous delivery is a development practice where the software functionality is deployed continuously to customer environment. Although this practice has been established in some domains such as B2C mobile software, the B2B domain imposes specific challenges. This article presents a case study that is conducted in a medium-sized software company operating in the B2B domain. The objective of this study is to analyze the challenges and benefits of continuous delivery in this domain. The results suggest that technical challenges are only one part of the challenges a company encounters in this transition. The company must also address challenges related to the customer and procedures. The core challenges are caused by having multiple customers with diverse environments and unique properties, whose business depends on the software product. Some customers require to perform manual acceptance testing, while some are reluctant towards new versions. By utilizing continuous delivery, it is possible for the case company to shorten the feedback cycles, increase the reliability of new versions, and reduce the amount of resources required for deploying and testing new releases.

\keywords {Continuous Delivery, Continuous Deployment, Development Process, B2B, Case Study}
\end{abstract}
%
\section{Introduction}
To deliver value fast and to cope with the increasingly active business environment, companies have to find solutions that improve efficiency and speed. Agile practices \cite{cockburn2002agile} have increased the ability of software companies to cope with changing customer requirements and changing market needs \cite{dzamashvili2010impact}. To even further increase the efficiency, shortening the feedback cycle enables faster customer feedback. Continuous delivery is a design practice that aims to shorten the delivery cycles by developing software in a way that it is always ready for releasing.

This study is an exploratory case study, which explores how continuous delivery can be applied in the case company that operates in the B2B domain. While existing studies of applying the practice exist \cite{olsson2012climbing, neely2013continuous}, none of the studies focuses specifically in the B2B domain. This study specifically aims to identify the main requirements, problems and key success factors with regards to continuous delivery in this domain. Extending the development process towards continuous delivery requires a deep analysis of the current development and deployment process, seeking the current problems and strengths. Adopting continuous delivery also requires understanding the requirements of continuous delivery, and restrictions caused by the developed software products.

@@

\section{Background and Related Work}
In the agile process software release is done in periodic intervals \cite{cockburn2002agile}. Compared to waterfall model it introduces multiple releases throughout the development. Continuous delivery, on the other hand, attempts to keep the software ready for release at all times during development process \cite{cdbook}. Continuous delivery is an extension to continuous integration, where the software functionality is kept in a state where it can be deployed to production immediately. Production deployments are manually triggered, but the entire deployment process is otherwise automated. While continuous integration defines a process where the work is automatically built, tested and frequently integrated to mainline \cite{duvall2007continuous}, often multiple times a day, continuous delivery adds automated acceptance testing and deployment to a staging environment. The purpose of continuous delivery is that as the deployment process is automated, it reduces human error and documents required for the build, and increases confidence that the build works \cite{cdbook}. It therefore aims to solve the problem of how to deliver an idea to users as quickly as possible.

Continuous delivery differs from continuous deployment. Continuous deployment means that every change goes through the pipeline and automatically gets put into production, resulting in many production deployments every day. Continuous delivery just means that you are able to do frequent deployments but may choose not to do it, usually due to businesses preferring a slower rate of deployment \cite{fowler}. An essential part of continuous delivery is the deployment pipeline, which is an automated implementation of an application’s build, deploy, test and release process \cite{humble2006deployment}. A deployment pipeline can be loosely defined as a consecutively executed set of validations that a software has to pass before it can be released. Common components of the deployment pipeline are a version control system and an automated test suite.

Challenges in adopting continuous delivery have been researched in multiple studies. Olsson et al. investigate the organization evolution path and the transition phase from continuous integration to continuous delivery \cite{olsson2012climbing}. The authors define continuous delivery as one of the final steps in the organization evolution path. The authors identify barriers that companies need to overcome to achieve the transition. One such barrier is the custom configuration at customer sites. Maintaining customized solutions and local configurations alongside the standard configurations creates issues. The second barrier is the internal verification loop, that has to be shortened not only to develop features faster but also to deploy fast. Finally, the lack of transparency and getting an overview of the status of development projects is seen as a barrier.

@@

@@

@@

\noindent Software development practices and product characteristics vary based on the domain and delivery model. Typical B2C applications are hosted as Software as a Service (SaaS) applications, and accessed by users via a web browser. In the B2B domain, applications installed to customer environments are very common. The purpose of this research question is to identify the challenges faced in applying continuous delivery in the B2B environment. The research question is answered by conducting interviews to discover the current development process and its challenges in the case company, and using these findings and the available literature on continuous delivery to map the initial set of challenges these approaches will encounter in the case company. The available literature is used to provide a thorough understanding of continuous delivery as a whole, so that challenges can be identified in all aspects of the practice.\newline

\noindent \textbf{RQ2: How does continuous delivery benefit the case company?} %What are the problems and strengths of the current development process?

@@

\subsubsection{Research design}
In this research, the units under the study are two teams within the case company, and the two software products developed by these teams. By focusing on two different products, a broader view on the application and consequences of the development approach can be gained. The first product, a marketing automation called Dialog, is used through an extensive user interface. The second product under inspection is a Master Data Management \cite{loshin2010master} solution running as an integrated background application.

%Methods
The primary source of information in this research are semi-structured interviews \cite{runeson2009guidelines}, performed within the two teams under study. The interview consists of pre-defined themes focusing on current development process, current deployment process, current interaction with customers, the software products and future ways with continuous delivery. Data is also collected through the product description documents and development process documents to verifying and supplementing the interview data.

@@

\section{Results}
This section is structured according to the research questions. The challenges regarding continuous delivery are analyzed in three areas: technical, procedural and customer. Technical aspect includes the environmental challenges, configural challenges and other challenges related to the software product and its usage. Procedural aspect includes the challenges regarding the software development process. Customer aspect consists of the customer interaction process and customer commitment.

@@

The developers perceive automated testing and test environments to be the largest technical task. The developers state that building a sufficient test automation is a very laborious process especially due to the maturity of the software, and are concerned with the maintainability of the test suite. The management is not sure what to test with automatic acceptance testing to validate a version.

@@

It is also common for B2B applications to have external components that have to be configured when the software is installed or the APIs to these components changed. The configurations for these external components either have to be manually updated, or automated as well. One of the main differences between B2B and B2C domains is the production environment. Both of the case company's products are used in multiple different customer environments. This introduces a problem of managing different configurations per customer environment and software instance.

@@

The basic deployment pipeline in the case company first includes a deploy to a user acceptance testing server, which is then tested manually by either the team or the customer. Only after the version has been acceptance tested and validated to work properly, can the production version be released. Continuous deployment to production is seen very risky due to the applications playing a major role in running the customers business.

Both of the case company's products are developed with a branching model, where feature branches are first thoroughly developed and then integrated to the master branch. With continuous delivery the long-lived feature branches should be changed to short-lived and relatively small feature branches to allow exposing new functionality faster to the customers, and receive feedback faster. While the small feature branches might be common for companies with a relatively new software products, companies that have been developing products for a long time might be more devoted to the practice of long-lived feature branches. %However, it is still possible to

The software applications in B2B often are large and modular applications, as is the case in the case company. The point when a deployment is triggered has to be designed to maintain the integrity of the application. As the deployment process is currently manually triggered by first releasing a version, a suitable time can be chosen each time. When a production deployment is triggered in continuous delivery, each module has to be in the correct state in order to produce a coherent version.

Both of the case company's products are used by multiple customers, each having their own environments. As the deployments are currently done manually, the customers receiving each deployment can be manually chosen.However, with a continuous delivery process whenever a feature or a new release is ready to be delivered, it can either be deployed to a single customer or to every customer.

%Versioning
Multiple customer environments affects versioning of the software product. In the case company, each customer has a unique configuration of the product, with possibly different versions of certain components. According to Jan Bosch, in an Innovation Experiment System environment only a single version exists: the currently deployed one. Other versions are retired and play no role \cite{bosch2012building}. However, with multiple environments, multiple different versions of the software are necessary at least in the early phase.

@@

Due to increased developer responsibility and varying interval of version updates continuous delivery causes, a team leader expresses concern that the delivery process complicates tracking when deployments are performed, and when features are finished. This also concerns other parties working in the customer interface, such as sales.

@@

Some customers of the case company are reluctant towards new releases. One of the reasons for this reluctancy is that new releases occasionally contain new bugs. In the case company, customers have been trained to perform certain tasks with a certain user interface. The customer might perform these tasks daily, once every two weeks or even less frequently. If the UI changes often, the customers feel lost and initially take more time to perform the tasks. This causes frustration in the users, and visible changes generally increases the reluctancy customers have towards new versions, unless the changes are significantly improving the user experience.  
\begin{aquote}{Product owner}
"The user interface should be easy to use. Now it's relatively hard to learn. If customers have just learned to perform a task, and we change the UI, the feedback is terrible."
\end{aquote}

Listing the changed features in changelog entries is especially important when releases are made more often. While the changes become smaller the faster versions are released, customers become less aware of when the version will be updated and when features have changed. Currently the version deployments are negotiated with the customers, and when the deployments are made more often, discussions regarding version releases may be reduced or even ceased.
